Marlowe: implementing and analysing financial contracts on blockchain

### 2019-12-09 papers/marlowe2019/paper.tex
First attempt at Marlowe intro.

Marlowe is a domain-specific language (DSL) for implementing financial contracts on blockchain: our initial target is Cardano, but it can be implemented on many DLT platforms, including Ethereum. Marlowe is embedded in Haskell, allowing users selectively to use aspects of Haskell -- typically definitions of constants and simple functions -- to express contracts more readably and succinctly. Marlowe was introduced in the paper~\cite{isola-marlowe}: Section \ref{sec:overview} below gives an overview of the language, and the changes made to it since the publication of that paper.

@@

Because Marlowe is a DSL, we are able to build special purpose tools and techniques to support it. Crucially in a financial environment we are able to \emph{exhaustively analyse contracts} without executing them, so that we can, for instance, check whether any particular contract makes all the payments it should: in the case it does not, we get an explicit example of how it can fail. This analysis is built into the Marlowe Playground. We are also able to use program verification techniques to prove guarantees about the implementation of Marlowe, including a guarantee that ``money in $=$ money out'' for all contracts.

@@

Since publication we have revised the language design, and so we give a brief overview of the current (3.0) version of the language and its informal semantics.\footnote{The complete Marlowe project, including semantics and implementation, is available at \url{https://github.com/input-output-hk/marlowe}.} Meadow has been upgraded to the Marlowe Playground, and that is also introduced here.

Introduce the main types and informal overview of execution. Contrast with Marlowe 1.3 (as in ISoLA paper). Embedding in Haskell? [All this stuff is in the tutorial already]

>>>>

Marlowe\footnote{The complete Marlowe project, including semantics and implementation, is available at \url{https://github.com/input-output-hk/marlowe}.} is a domain-specific language (DSL) for implementing financial contracts on blockchain: our initial target is Cardano, but it can be implemented on many DLT platforms, including Ethereum. Marlowe is embedded in Haskell, allowing users selectively to use aspects of Haskell -- typically definitions of constants and simple functions -- to express contracts more readably and succinctly. Marlowe was introduced in the paper~\cite{isola-marlowe}: Section \ref{sec:overview} below gives an overview of the language, and the changes made to it since the publication of that paper.

@@

Because Marlowe is a DSL, we are able to build special purpose tools and techniques to support it. Crucially in a financial environment we are able to \emph{exhaustively analyse contracts} without executing them, so that we can, for instance, check whether any particular contract makes all the payments it should: in the case it does not, we get an explicit example of how it can fail. This analysis, reported in Section \ref{sec:analysis}, is built into the Marlowe Playground. This section also covers our use of formal verification to prove properties of the implementation of Marlowe, including a guarantee that ``money in $=$ money out'' for all contracts.

@@

Since first publication we have revised the language design: this section gives a brief overview of the current (3.0) version of the language and its semantics.

\subsubsection*{The Marlowe model}

Marlowe is designed to support the execution of financial contracts on blockchain, and specifically to work on Cardano. Contracts are built by putting together a small number of constructs that in combination can be used to describe many different kinds of financial contract.

Contracts in Marlowe run on a blockchain, but need to interact with the off-chain world. The parties to the contract, also called the participants, can engage in various actions: they can be asked to deposit money, or to make a choice between various alternatives. A notification of an external / oracle value is the other possible form of input. Running a contract will also produce external effects, by making payments to parties in the contract.

The Marlowe model allows for a contract to control money in a number of disjoint accounts: this allows for more explicit control of how the money flows in the contract. Each account is owned by a particular party to the contract, and that party receives a refund of any remaining funds in the account when the contract is closed. %These accounts are local, in that they only exist as during the execution of the contract, and during that time they are only accessible by parties to the contract.

Marlowe contracts describe a series of steps, typically by describing the first step, together with another (sub-) contract that describes what to do next. For example, the contract \texttt{Pay a p v cont} says ``make a payment of \texttt{v} Lovelace to the party \texttt{p} from the account \texttt{a}, and then follow the contract \texttt{cont}''. We call \texttt{cont} the continuation of the contract.

In executing a contract we need to keep track of the current contract: after making a step in the example above, the current contract is \texttt{cont}. We also have to keep track of some other information, such as how much is held in each account: this information together is the state, and generally this changes at each step. A step can also see an action taking place, such as money being deposited, or an effect being produced, e.g. a payment.
%Value on the blockchain resides in the UTxO, which are protected cryptographically by a private key held by the owner. These keys can be used to redeem the output, and so to use them as inputs to new transactions, which can be seen as spending the value in the inputs. Users typically keep track of their private keys, and the values attached to them, in a cryptographically-secure wallet.
It is through their wallets that users are able to interact with Marlowe contracts running on the blockchain, making deposits and receiving payments. %Note, however, that these are definitely off chain actions that need to be controlled by code running in the user's wallet: they cannot be made to happen by the Marlowe contract itself.

%\subsection{Executing a Marlowe contract}
%At each slot, a running Marlowe contract will receive a list of inputs in order. The contract is executed by evaluating it step by step until it cannot be changed any further without processing any input, a condition that we call being quiescent. The first input is then processed, and then the contract is single stepped again until quiescence, and this process is repeated until all the inputs are processed. At each step the current contact and the state will change, some input may be processed, and payments made.

\subsubsection*{Marlowe step by step}
Marlowe has five ways of building contracts. Four of these -- \texttt{Pay}, \texttt{Let}, \texttt{If} and \texttt{When} -- build a complex contract from simpler contracts, and the fifth, \texttt{Close}, is a simple contract. At each step of execution, as well as returning a new state and continuation contract, it is possible that effects, e.g.\ payments, and warnings can be generated too. We also explain Marlowe values, observations and actions, which are used to supply external information and inputs to a running contract to control how it will evolve.

\paragraph{Pay}
A payment contract \texttt{Pay a p v} cont will make a payment of value \texttt{v} from the account \texttt{a} to a payee \texttt{p}, which will be one of the contract participants or another account in the contract. Warnings will be generated if the value \texttt{v} is negative, or if there is not enough in the account to make the payment in full. In that case a partial payment (of all the money available) is made. The continuation contract is the one given in the contract: \texttt{cont}.

\paragraph{Close}
A contract \texttt{Close} provides for the contract to be closed (or terminated). The only action that is performs is to provide refunds to the owners of accounts that contain a positive balance. This is performed one account per step, but all accounts will be refunded in a single transaction.

%Before discussing other forms of contracts, we need to describe values, observations and actions.

\paragraph{Values, observations and actions}
\emph{Values} include some quantities that change with time, including ``the current slot number'', ``the current balance of an account (in Lovelace)'', and any choices that have already been made; we call these volatile values. Values can also be combined using addition, subtraction and negation.

\emph{Observations} are Boolean values derived by comparing values, and can be combined using the standard Boolean operators. It is also possible to observe whether any choice has been made (for a particular identified choice).

Observations will have a value at every step of execution. On the other hand, \emph{actions} happen at particular points during execution and can be
(i) depositing money, (ii)
making a choice between various alternatives, or (iii)
notifying an external or oracle value, such as the current price of a particular commodity.

\paragraph{If}
The conditional \texttt{If obs cont1 cont2} will continue as \texttt{cont1} or \texttt{cont2}, depending on the Boolean value of the observation \texttt{obs} on execution.

\paragraph{When}
This is the most complex constructor for contracts, with the form \texttt{When cases timeout cont}. It is a contract that is triggered on actions, which may or may not happen at any particular slot: what happens when various actions happen is described by the cases in the contract.

In the contract \texttt{When cases timeout cont}, the list \texttt{cases} contains a collection of cases. Each case has the form \texttt{Case ac co} where \texttt{ac} is an action and \texttt{co} a continuation (another contract). When a particular action, e.g.  \texttt{ac}, happens, the state is updated accordingly and the contract will continue as the corresponding continuation \texttt{co}.

In order to make sure that the contract makes progress eventually, the contract \texttt{When cases timeout cont} will continue as cont once the \texttt{timeout}, a slot number, is reached.

\paragraph{Let}
A let contract \texttt{Let id val cont} causes the expression \texttt{val} to be evaluated, and stored with the name \texttt{id}. The contract then continues as \texttt{cont}.

%As well as allowing us to use abbreviations, this mechanism also means that we can capture and save volatile values that might be changing with time, e.g. the current price of oil, or the current slot number, at a particular point in the execution of the contract, to be used later on in contract execution.

%%% end overview
